Dynamical modeling of tidal streams

### 2013-12-31 tex/ms.tex
streams: text tweaks after first perusal

\newcommand{\jr}{\ensuremath{J_R}}
\newcommand{\jphi}{\ensuremath{J_\phi}}
\newcommand{\jz}{\ensuremath{J_Z}}
\newcommand{\lz}{\ensuremath{L_Z}}
\newcommand{\Or}{\ensuremath{\Omega_R}}
\newcommand{\Ophi}{\ensuremath{\Omega_\phi}}
\newcommand{\Oz}{\ensuremath{\Omega_Z}}
\newcommand{\ar}{\ensuremath{\theta_R}}
\newcommand{\aphi}{\ensuremath{\theta_\phi}}
\newcommand{\az}{\ensuremath{\theta_Z}}
\newcommand{\apar}{\ensuremath{\theta_\parallel}}
\newcommand{\opar}{\ensuremath{\Omega_\parallel}}
\newcommand{\aperp}{\ensuremath{\veca_\perp}}
\newcommand{\aperpii}{\ensuremath{\theta_{\perp,ii}}}
\newcommand{\operp}{\ensuremath{\veco_\perp}}
\newcommand{\operpii}{\ensuremath{\Omega_{\perp,ii}}}

@@

\begin{abstract} 
  I present a new framework for modeling the dynamics of cold tidal
  streams. The framework consists of simple models for the initial
  action--angle distribution of tidal debris as a function of time,
  which can be straightforwardly evolved forward in time. Taking
  advantage of the essentially one-dimensional nature of cold tidal
  streams, the transformation to position--velocity coordinates can be
  linearized and interpolated near a small number of points along the
  progenitor orbit, thus allowing for efficient computations of a
  stream's properties in observable quantities. I illustrate how to
  calculate the stream's track in different coordinate systems, how to
  quickly estimate the dispersion perpendicular to the track, and how
  to draw mock stream data. As a generative model, this framework
  allows one to compute the full probability distribution function and
  marginalize over or condition it on certain phase--space dimensions
  as well as convolve it with observational uncertainties. This will
  be instrumental in proper data analysis of stream data. In addition
  to providing a computationally-efficient practical tool for modeling
  the dynamics of tidal streams, the action--angle nature of the
  framework helps elucidate in exactly what manner streams do not
  follow single orbits, how the observed width of the stream relates
  to the velocity dispersion or mass of the progenitor, and how the
  progenitor of ``orphan'' streams could be located.

@@

It has long been clear that the dynamics of tidal streams is most
simply described in terms of action--angle coordinates
\citep{Tremaine99a,Helmi99a}. Once a star has been tidally stripped
from the progenitor, the self-gravity of the stream can be neglected
and the orbital actions of a stream member are conserved while the
angles increase linearly with time. The action--angle structure of a
stream is therefore characterized by strong but straightforward
correlations between the actions and angles of stream members, the
exploitation of which is crucial for using streams to measure the host
gravitational potential \citep{Sanders13b}. The description of a
stream in action--angle coordinates also elucidates the connection
between the orbit and velocity dispersion (or mass) of the progenitor
and the action distribution of the tidal debris \citep{Eyre11a},
allowing simple physical models of the stream to be used and
constrained by observational data.

@@

In this paper, I present a new method for modeling cold tidal streams
that fundamentally lives in frequency--angle space, because after
specifying the time at which a star was stripped, the stream
distribution function is essentially one-dimensional in this
space. While many of the results of this paper are more generally
valid, the fiducial stream model in this paper consists of a
three-dimensional (close to) Gaussian distribution of frequencies in
the stream, a three-dimensional Gaussian distribution of initial angle
offsets between stream members and the progenitor, and a uniform
distribution of stripping times; the frequency and angle distributions
are independent of stripping time. While this model oversimplifies the
real stripping process, it is argued that at least for constraining
the gravitational potential with stream data, this simple model is
likely unbiased. For detailed mock stream data, more elaborate models
of the stripping process as a function of time might be necessary and
I discuss how these could easily be incorporated into the proposed
framework.

@@

In real space, a tidal stream forms as stars are stripped from a
progenitor some time $\Delta t$ in the past and then evolve (largely)
independently in a galaxy's host potential, moving away both in
position and velocity from the progenitor object as time goes
on. Thus, a star was offset by a small amount $(\Delta \vecx,\Delta
\vecv)$ from the position of the progenitor $(\vecx^p,\vecv^p)(-\Delta
t)$ at that time. Then the star orbited under the influence of the
same gravitational potential $\Phi$ as the progenitor until the
present day $(t=0)$, when it was observed: $(\vecx,\vecv)(t=0)$ =
$\vech_{\Delta t}(\vecx^p(-\Delta t)+\Delta \vecx,\vecv^p(-\Delta
t)+\Delta \vecv)$, where $\vech$ denotes the Hamiltonian flow from
$t=-\Delta t$ to $t=0$ (see \citealt{binneytremaine}); this is simply
the orbit of the star between these times. In phase-space this orbit
can be calculated by solving Hamilton's equations: $\dot{\vecx} =
\vecv; \dot{\vecv} = - \dd \Phi / \dd \vecx$.

In action--angle space the dynamics of stream formation is much
simpler. A star received an offset ($\Delta \vecj,\Delta \veca$) at
time $-\Delta t$ from the action--angle coordinates of the progenitor
$(\vecj^p,\veca^p)$, similar to the small offset ($\Delta \vecx,\Delta
\vecv)$ in the real--space description. The offset in the actions
corresponds to an offset in the frequencies $\Delta \veco$;
ultimately, this frequency offset is responsible for the spreading of
the stream over long stretches on the sky. Because both the
progenitor's angles and the star's angles increase linearly with time,
albeit at different frequencies, the difference in the angles also
increases linearly in time. Therefore at time $t=0$ the difference in
angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\Delta t) + \Delta \veco \Delta t\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\Delta t)\,.
\end{equation}
As the initial offset $\Delta \veca(-\Delta t)$ in angles is small
compared to the growth, we have that $\Delta\veco\Delta t$, $\Delta
\veca(t=0) \approx \Delta \veco \Delta t$.

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial \veco /
\partial \vecj) \Delta \vecj = (\partial^2 H / \partial \vecj \partial
\vecj) \Delta \vecj$, in the limit of small action offsets, because
$\veco = \partial H / \partial \vecj$. For an approximately
one-dimensional stream to form, the Hessian matrix $(\partial^2 H /
\partial \vecj \partial \vecj)\large|_{\vecj^p}$ has to be dominated
by a single large eigenvalue; in this case the angle difference of
stars stripped from the stream at any time will fall along
approximately the same direction $\unitvec\veco$, and the stream will
essentially be one dimensional. In detail, because the distribution of
initial offsets $\Delta \vecj$ is not isotropic, the direction
$\unitvec \veco$ is not just determined by the potential (through $H$)
and the progenitor orbit, but also by the distribution of $\Delta
\vecj$, which is set through a combination of the internal dynamics of
the progenitor and the stripping process (see discussion in
\citealt{Sanders13b}). For dynamical modeling we can be agnostic about
the relative importance of all of these contributors to $\unitvec
\veco$ by flexible modeling of the distribution of $\Delta \vecj$ or
$\Delta \veco$.

@@

I illustrate the theoretical considerations set forth in this paper by
using a simulated tidal stream. This stream is set up similarly to
that used in \citet{Sanders13b}. The stream is constructed by running
an $N$-body simulation of a King cluster on an orbit similar to that
of the GD-1 stream \citep{Koposov10a}. Specifically, a King cluster
with a mass of $2\times10^4\msun$, a tidal radius of $0.07\kpc$, and a
ratio of the central potential to the velocity dispersion squared of
$2.0$ is sampled using $10^4$ particles. The cluster is evolved in an
external logarithmic host potential with $V_c = 220\kms$ and a
flattening $q=0.9$ ($\Phi = V_c^2 \ln [R^2 + z^2/q^2]$) using the
gyrfalcon code \citep{Dehnen00a,Dehnen02a} with a softening of
$1.5\pc$ in the NEMO toolkit \citep{Teuben95a}. The cluster is evolved
for $5.011\Gyr$ ($=5.125\kpc/[\kms]$) from the initial condition
$(X,Y,Z) = (-11.63337239,-20.76235661,-10.631736273934635)\kpc$ and
$(V_X,V_Y,V_Z) = (-128.8281653,42.88727925,79.172383882274971)\kms$,
chosen such that the cluster ends up at roughly the observed location
of the GD-1 stream at the end of the simulation
\citep{Koposov10a}. The progenitor's orbit has an eccentricity of
$0.31$, a pericenter of $13.7\kpc$, a current $r=14.4\kpc$, such that
the cluster is close to pericenter at the end of the simulation. The
actions of the progenitor are $(J_R,L_Z,J_Z) =
(288.5,-3173.7,897.6)\kpc\kms$ and its orbital frequencies are
$(\Omega_R,\Omega_\phi,\Omega_Z) = (15.7,-10.8,11.9)\Gyr\inv$.

@@

The action and frequency distributions shown in
\figurename~\ref{fig:gd1_jt} combined with the knowledge that for
\emph{any} tidal stream the Hessian $(\partial^2 H / \partial \vecj
\partial \vecj)\large|_{\vecj^p}$ must be strongly dominated by a
single large eigenvalue leads us to propose that frequency--angle
space is a better coordinate system than action--angle space for
modeling tidal streams. That is, the distribution of actions in a
tidal stream has a complicated structure that is hard to capture using
a simple, analytic form (see also the discussion in
\citealt{Eyre11a}). Especially the ``bowtie'' structure in $(J_R,L_Z)$
and the overlapping leading- and trailing-arm distributions in
($J_R,J_Z$) are hard to model with a simple distribution function. The
distribution of frequencies, however, is close-to one dimensional, its
direction of largest variance can be well-modeled as a Gaussian
(although below we will model it slightly differently), and its
parameters can all be easily estimated from the velocity dispersion of
the progenitor, the orbit of the progenitor, and the gravitational
potential, as discussed below. Therefore, we will model tidal streams
in frequency--angle space. Because the leading and trailing arm are
well-separated in frequency space, we model each arm individually.

@@

Ingredients (a) and (b) require more careful modeling. The right panel
of \figurename~\ref{fig:gd1_times} shows the distribution of times at
which stream particles are stripped from the progenitor. These
stripping times are calculated from the final snapshot by linear
regression of a particle's angle offset $\Delta \veca$ from the
progenitor, versus the frequency offset $\Delta \veco$ as 
\begin{equation}
  \Delta t = \frac{\Delta \veco \cdot \Delta \veca}{|\Delta \veco|^2}\,.
\end{equation}
The dotted line shows a finely-binned histogram of these stripping
times, which indicates that stripping happens in bursts with a radial
period of about $400\Myr$, the radial period of the progenitor's
orbit. Thus, as expected, stripping happens primarily at pericenter
passages, with a smaller number of particles lost between pericenter
passages. Realistic models of a stream, especially those that require
good models of the surface-density structure of the stream as a
function of stream angle, need to model this non-uniform stripping
process.

@@

The right panel of \figurename~\ref{fig:gd1_times} shows the angle
offset (at the end of the simulation) versus the stripping time. There
is a spread in the angle offset reach for any given stripping time,
but it is nevertheless typically the case that particles that have
been stripped earlier are found at larger angle offsets
today. Self-sorting---the tendency of particles stripped later with
larger frequency offsets to overtake particles stripped earlier at
smaller frequency differences and thus to erase correlations between
particle position along the stream and stripping time---happens to a
small extent but the narrow distribution of frequency offsets limits
the ability of particles stripped later to reach large distances from
the progenitor.

@@

First, we approximate the bursty, non-uniform distribution of
stripping times in \figurename~\ref{fig:gd1_times} with a uniform
distribution
\begin{align}\label{eq:pt}
  \begin{split}
    p(\Delta t) & \propto 1\,,\qquad 0 < \Delta t < t_d\\
    & = 0\,,\qquad \mathrm{otherwise}\,.
  \end{split}
\end{align}
The thick, solid histogram in the right panel of
\figurename~\ref{fig:gd1_times} shows a coarser-binned histogram of
stripping times, indicating that the uniform distribution is a good
approximation over intervals larger than a radial period ($\approx
400\Myr$). The \emph{disruption time} $t_d$ is a free parameter of the
model that needs to be determined for each stream. For the model of
the simulated stream used here we set $t_d = 4.5\Gyr$.

@@

We then further constrain the mean frequency offset to lie along the
principal eigenvector of $V_\veco$. This mean frequency offset can be
described in terms of a parameter $\mu_\Omega = \Delta \Omega^m /
\sigma_{\Omega,1}$, where $\sigma_{\Omega,1}$ is the square root of
the largest eigenvalue of $V_\veco$ and $\Delta \Omega^m$ is the mean
frequency offset. The sign of this parameter sets whether we are
modeling a leading or a trailing arm. \figurename~\ref{fig:gd1_dohist}
shows the distribution frequency offsets in the direction of the
median frequency offset in the stream. The dashed line is a Gaussian
fit to this distribution, while the solid line is a fit of the form
\begin{equation}\label{eq:do}
  p(|\Delta \veco|_\parallel) = |\Delta \veco|_\parallel\,\mathcal{N}\left(|\Delta \veco|_\parallel|\Delta \Omega^m,\sigma_{\Omega,1}^2\right)\,,
\end{equation}
where $\mathcal{N}(\cdot|\cdot,\cdot)$ represents a Gaussian
distribution. The fit is equally good (because the dispersion is much
smaller than the mean) and we choose the second form because it
simplifies some of the stream-track calculations below. The best-fit
$(\Delta \Omega^m,\sigma_{\Omega,1})$ is $(0.19,0.033)\Gyr\inv$ in
both cases. Based on this information, a model with $\sigv =
0.365\kms$ and $\mu_\Omega = 6$ provides a good fit to the simulation
data. The distribution of frequency offsets in the two-dimensional
space perpendicular to the principal eigenvector of $V_\veco$ is
modeled as a zero-mean Gaussian with variance matrix given by the
projection of $V_\veco$ onto this space.

Because we expect the typical $\Delta\veco$ to scale with the
velocity dispersion of the progenitor (see also \citealt{Sanders13a}),
I conjecture that a constant $\mu_\Omega$ can be used for modeling
cold tidal streams of any (small-ish) velocity dispersion. Further
modeling of simulated streams, however, is necessary to check this and
to find the optimizal value of $\mu_\Omega$. Fixing $\mu_\Omega$
leaves a single free parameter \sigv\ for modeling the frequency
offset distribution of a tidal stream. In modeling to observed data,
this parameter needs to be fit and we expect it to be proportional to
the velocity dispersion of the progenitor, although this
proportionality needs to be checked more carefully (but see Figure 3
in \citealt{Sanders13a}, which shows that the size of the frequency
distribution scales as mass$^{1/3}$, or approximately as the velocity
dispersion). Similarly, the initial angle distribution, modeled here
as an isotropic Gaussian, has a charactaristic spread $\sigma_{\theta}$
which we set here to $0.00365$ ($=\sigv/[100\kms]$), and we likewise
expect this spread to scale with the velocity dispersion of the
progenitor.

The fiducial model for a tidal stream in frequency--angle coordinates
he proposed here is therefore described by essentially two free
parameters in addition to the progenitor's phase-space position: the
progenitor's velocity dispersion \sigv\ and the disruption time
$t_d$. As shown below, these two parameters are the most important in
determining the track of the stream. More sophisticated models can
leave $\mu_\Omega$ and $\sigma_{\theta}$ as free paramters as well. In
what follows, we will fix these to the values given in the previous
paragraph. It must be stressed, however, that all of these parameters
here have been fixed ``by-eye'', and better fits might be possible
using more quantitative fit procedures. But such procedures are not
pursued here further.

@@

Here, I discuss how to calculate the stream's properties as a function
of the angle along the stream. In the generative model described
above, the direction of the angle along the stream lies along the
principal eigenvector of $V_\veco$, the variance matrix of the model
$\Delta \veco$ distribution. We denote the angle in this direction as
$\apar$ and the frequency in this direction $\opar$; the
two-dimensional angle- and frequency-space perpendicular to this
direction is denoted as $\aperp$ and $\operp$.

>>>>

\newcommand{\jr}{\ensuremath{J_R}}
\newcommand{\jphi}{\ensuremath{J_\phi}}
\newcommand{\jz}{\ensuremath{J_Z}}
\newcommand{\lz}{\ensuremath{L_Z}}
\newcommand{\Or}{\ensuremath{\Omega_R}}
\newcommand{\Ophi}{\ensuremath{\Omega_\phi}}
\newcommand{\Oz}{\ensuremath{\Omega_Z}}
\newcommand{\ar}{\ensuremath{\theta_R}}
\newcommand{\aphi}{\ensuremath{\theta_\phi}}
\newcommand{\az}{\ensuremath{\theta_Z}}
\newcommand{\apar}{\ensuremath{\theta_\parallel}}
\newcommand{\opar}{\ensuremath{\Omega_\parallel}}
\newcommand{\aperp}{\ensuremath{\veca_\perp}}
\newcommand{\aperpii}{\ensuremath{\theta_{\perp,ii}}}
\newcommand{\operp}{\ensuremath{\veco_\perp}}
\newcommand{\operpii}{\ensuremath{\Omega_{\perp,ii}}}
\newcommand{\ts}{\ensuremath{t_s}}

@@

\begin{abstract} 
  I present a new framework for modeling the dynamics of cold tidal
  streams. The framework consists of simple models for the initial
  action--angle distribution of tidal debris, which can be
  straightforwardly evolved forward in time. Taking advantage of the
  essentially one-dimensional nature of cold tidal streams, the
  transformation to position--velocity coordinates can be linearized
  and interpolated near a small number of points along the progenitor
  orbit, thus allowing for efficient computations of a stream's
  properties in observable quantities. I illustrate how to calculate
  the stream's average location (its ``track'') in different
  coordinate systems, how to quickly estimate the dispersion around
  its track, and how to draw mock stream data. As a generative model,
  this framework allows one to compute the full probability
  distribution function and marginalize over or condition it on
  certain phase--space dimensions as well as convolve it with
  observational uncertainties. This will be instrumental in proper
  data analysis of stream data. In addition to providing a
  computationally-efficient practical tool for modeling the dynamics
  of tidal streams, the action--angle nature of the framework helps
  elucidate in exactly what manner streams do not follow single
  orbits, how the observed width of the stream relates to the velocity
  dispersion or mass of the progenitor, and how the progenitors of
  ``orphan'' streams could be located.

@@

It has long been clear that the dynamics of tidal streams is most
simply described in terms of action--angle coordinates
\citep{Tremaine99a,Helmi99a}. Once a star has been tidally stripped
from the progenitor, the self-gravity of the stream can be neglected
and the orbital actions of a stream member are conserved while the
angles increase linearly with time. The action--angle structure of a
stream is therefore characterized by strong but straightforward
correlations between the actions and angles of stream members, the
exploitation of which is crucial for using streams to measure the host
gravitational potential \citep{Sanders13b}. The description of a
stream in action--angle coordinates also elucidates the connection
between the orbit and velocity dispersion (or mass) of the progenitor
and the action distribution of the tidal debris \citep{Eyre11a},
allowing simple physical models of the stream to be used and to be
constrained by observational data.

@@

In this paper, I present a new method for modeling cold tidal streams
that fundamentally lives in frequency--angle space, because the stream
distribution function is essentially one-dimensional in this
space. While many of the results of this paper are more generally
valid, the fiducial stream model in this paper consists of a
three-dimensional (close to) Gaussian distribution of frequencies in
the stream, a three-dimensional Gaussian distribution of initial angle
offsets between stream members and the progenitor, and a uniform
distribution of stripping times; the frequency and angle distributions
are independent of stripping time. While this model oversimplifies the
real stripping process, it is argued that at least for constraining
the gravitational potential with stream data, this simple model is
likely unbiased. For detailed mock stream data, more elaborate models
of the stripping process as a function of time might be necessary and
I discuss how these could easily be incorporated into the proposed
framework.

@@

In real space, a tidal stream forms as stars are stripped from a
progenitor cluster or satellite galaxy some time $\ts$ in the
past and then evolve (largely) independently in a galaxy's host
potential, moving away both in position and velocity from the
progenitor object as time goes on. Thus, a star was offset by a small
amount $(\Delta \vecx,\Delta \vecv)$ from the position of the
progenitor $(\vecx^p,\vecv^p)(-\ts)$ at that time. Then the star
orbited under the influence of the same gravitational potential $\Phi$
as the progenitor until the present day $(t=0)$, when it was observed:
$(\vecx,\vecv)(t=0)$ = $\vech_{\ts}(\vecx^p(-ts)+\Delta
\vecx,\vecv^p(-\ts)+\Delta \vecv)$, where $\vech$ denotes the
Hamiltonian flow from $t=-\ts$ to $t=0$ (see
\citealt{binneytremaine}); this is simply the orbit of the star
between these times. In phase-space this orbit can be calculated by
solving Hamilton's equations: $\dot{\vecx} = \vecv; \dot{\vecv} = -
\dd \Phi / \dd \vecx$.

In action--angle space the dynamics of stream formation is much
simpler. A star received an offset ($\Delta \vecj,\Delta \veca$) at
time $-\ts$ from the action--angle coordinates of the progenitor
$(\vecj^p,\veca^p)$, similar to the small offset ($\Delta \vecx,\Delta
\vecv)$ in the real--space description. The offset in the actions
corresponds to an offset in the frequencies $\Delta \veco$;
ultimately, this frequency offset is responsible for the spreading of
the stream over long stretches on the sky. Because both the
progenitor's angles and the star's angles increase linearly with time,
albeit at different frequencies, the difference in the angles also
increases linearly in time. Therefore at time $t=0$ the difference in
angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\ts) + \Delta \veco \,\ts\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\ts)\,.
\end{equation}
As the initial offset $\Delta \veca(-\ts)$ in angles is small compared
to the growth $\Delta \veco \,\ts$, we have that $\Delta \veca(t=0)
\approx \Delta \veco \,\ts$.

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial \veco /
\partial \vecj) \Delta \vecj = (\partial^2 H / \partial \vecj \partial
\vecj) \Delta \vecj$, in the limit of small action offsets, because
$\veco = \partial H / \partial \vecj$, where $H$ is the
Hamiltonian. For an approximately one-dimensional stream to form, the
Hessian matrix $(\partial^2 H / \partial \vecj \partial
\vecj)\large|_{\vecj^p}$ has to be dominated by a single large
eigenvalue; in this case the angle difference of stars stripped from
the stream at any time will fall along approximately the same
direction $\unitvec\veco$, and the stream will be essentially one
dimensional. In detail, because the distribution of initial offsets
$\Delta \vecj$ is not isotropic, the direction $\unitvec \veco$ is not
just determined by the potential (through $H$) and the progenitor
orbit, but also by the distribution of $\Delta \vecj$, which is set
through a combination of the internal dynamics of the progenitor and
the stripping process (see discussion in \citealt{Sanders13b}). For
dynamical modeling we can be agnostic about the relative importance of
all of these contributors to $\unitvec \veco$ by flexible modeling of
the distribution of $\Delta \vecj$ or $\Delta \veco$.

@@

I illustrate the theoretical considerations set forth in this paper by
using a simulated tidal stream. This stream is set up similarly to
that used in \citet{Sanders13b}. The stream is constructed by running
an $N$-body simulation of a King cluster on an orbit similar to that
of the GD-1 stream \citep{Koposov10a}. Specifically, a King cluster
with a mass of $2\times10^4\msun$, a tidal radius of $0.07\kpc$, and a
ratio of the central potential to the velocity dispersion squared of
$2.0$ is sampled using $10^4$ particles. The cluster is evolved in an
external logarithmic host potential with $V_c = 220\kms$ and a
flattening $q=0.9$ ($\Phi = V_c^2 \ln [R^2 + z^2/q^2]$) using the
gyrfalcon code \citep{Dehnen00a,Dehnen02a} with a softening of
$1.5\pc$ in the NEMO toolkit \citep{Teuben95a}. The cluster is evolved
for $5.011\Gyr$ ($=5.125\kpc/[\kms]$) from the initial condition
$(X,Y,Z) = (-11.63337239,-20.76235661,-10.631736273934635)\kpc$ and
$(V_X,V_Y,V_Z) = (-128.8281653,42.88727925,79.172383882274971)\kms$,
chosen such that the cluster ends up at roughly the observed location
of the GD-1 stream at the end of the simulation
\citep{Koposov10a}. The progenitor cluster's orbit has an eccentricity
of $0.31$, a pericenter of $13.7\kpc$, and a current $r=14.4\kpc$,
such that the cluster is close to pericenter at the end of the
simulation. The actions of the progenitor are $(J_R,L_Z,J_Z) =
(288.5,-3173.7,897.6)\kpc\kms$ and its orbital frequencies are
$(\Omega_R,\Omega_\phi,\Omega_Z) = (15.7,-10.8,11.9)\Gyr\inv$.

@@

The action and frequency distributions shown in
\figurename~\ref{fig:gd1_jt} combined with the knowledge that for
\emph{any} tidal stream the Hessian $(\partial^2 H / \partial \vecj
\partial \vecj)\large|_{\vecj^p}$ must be strongly dominated by a
single large eigenvalue leads me to propose that frequency--angle
space is a better coordinate system than action--angle space for
modeling tidal streams. That is, the distribution of actions in a
tidal stream has a complicated structure that is hard to capture using
a simple, analytic form (see also the discussion in
\citealt{Eyre11a}). Especially the ``bowtie'' structure in $(J_R,L_Z)$
and the overlapping leading- and trailing-arm distributions in
($J_R,J_Z$) are difficult to model with a simple distribution
function. The distribution of frequencies, however, is close-to one
dimensional, its direction of largest variance can be well-modeled as
a Gaussian (although below we will model it slightly differently), and
its parameters can all be easily estimated from the velocity
dispersion of the progenitor, the orbit of the progenitor, and the
gravitational potential, as discussed below. Therefore, we will model
tidal streams in frequency--angle space. Because the leading and
trailing arm are well-separated in frequency space, we model each arm
individually.

@@

Ingredients (a) and (b) require more careful modeling. The right panel
of \figurename~\ref{fig:gd1_times} shows the distribution of times at
which stream particles are stripped from the progenitor. These
stripping times are calculated from the final snapshot by linear
regression of a particle's angle offset $\Delta \veca$ from the
progenitor, versus the frequency offset $\Delta \veco$ as 
\begin{equation}
  \ts = \frac{\Delta \veco \cdot \Delta \veca}{|\Delta \veco|^2}\,.
\end{equation}
The dotted line shows a finely-binned histogram of these stripping
times, which indicates that stripping happens in bursts with a period
of about $400\Myr$, the radial period of the progenitor's orbit. Thus,
as expected, stripping happens primarily at pericenter passages, with
a smaller number of particles lost between pericenter
passages. Realistic models of a stream, especially those that require
good models of the surface-density structure of the stream as a
function of stream angle, need to model this non-uniform stripping
process.

@@

The right panel of \figurename~\ref{fig:gd1_times} shows the angle
offset (at the end of the simulation) versus the stripping time. There
is a spread in the angle offset reach for any given stripping time,
but it is nevertheless typically the case that particles that have
been stripped earlier are found at larger angle offsets
today. Self-sorting---the tendency of particles stripped later with
larger frequency offsets to overtake particles stripped earlier at
smaller frequency differences and thus to erase correlations between
particle position along the stream and stripping time---happens to a
small extent but the narrow distribution of frequency offsets limits
the ability of particles removed later to reach large distances from
the progenitor.

@@

First, we approximate the bursty, non-uniform distribution of
stripping times in \figurename~\ref{fig:gd1_times} with a uniform
distribution
\begin{align}\label{eq:pt}
  \begin{split}
    p(\ts) & \propto 1\,,\qquad 0 < \ts < t_d\\
    & = 0\,,\qquad \mathrm{otherwise}\,.
  \end{split}
\end{align}
The thick, solid histogram in the right panel of
\figurename~\ref{fig:gd1_times} shows a coarser-binned histogram of
stripping times, indicating that the uniform distribution is a good
approximation over intervals larger than a radial period ($\approx
400\Myr$). The \emph{disruption time} $t_d$ is a free parameter of the
model that needs to be determined for each stream. For the model of
the simulated stream used here we set $t_d = 4.5\Gyr$.

@@

We then further constrain the mean frequency offset to lie along the
principal eigenvector of $V_\veco$. This mean frequency offset can be
described in terms of a parameter $\mu_\Omega = \Delta \Omega^m /
\sigma_{\Omega,1}$, where $\sigma_{\Omega,1}$ is the square root of
the largest eigenvalue of $V_\veco$ and $\Delta \Omega^m$ is the
one-dimensional mean frequency offset. The sign of this parameter sets
whether we are modeling a leading or a trailing
arm. \figurename~\ref{fig:gd1_dohist} shows the distribution of
frequency offsets in the direction of the median frequency offset in
the stream. The dashed line is a Gaussian fit to this distribution,
while the solid line is a fit of the form
\begin{equation}\label{eq:do}
  p(|\Delta \veco|_\parallel) = |\Delta \veco|_\parallel\,\mathcal{N}\left(|\Delta \veco|_\parallel|\Delta \Omega^m,\sigma_{\Omega,1}^2\right)\,,
\end{equation}
where $\mathcal{N}(\cdot|\cdot,\cdot)$ represents a Gaussian
distribution. The fit is equally good (because the dispersion is much
smaller than the mean) and we choose the second form because it
simplifies some of the stream-track calculations below. The best-fit
$(\Delta \Omega^m,\sigma_{\Omega,1})$ is $(0.19,0.033)\Gyr\inv$ in
both cases. Based on this information, a model with $\sigv =
0.365\kms$ and $\mu_\Omega = 6$ provides a good fit to the simulation
data. The distribution of frequency offsets in the two-dimensional
space perpendicular to the principal eigenvector of $V_\veco$ is
modeled as a zero-mean Gaussian with variance matrix given by the
projection of $V_\veco$ onto this space.

Because we expect the typical $\Delta\veco$ to scale with the
velocity dispersion of the progenitor (see also \citealt{Sanders13a}),
I conjecture that a constant $\mu_\Omega$ can be used for modeling
cold tidal streams of any (small-ish) velocity dispersion. Further
modeling of simulated streams, however, is necessary to check this and
to find the optimizal value of $\mu_\Omega$. Fixing $\mu_\Omega$
leaves a single free parameter \sigv\ for modeling the frequency
offset distribution of a tidal stream. In modeling to observed data,
this parameter needs to be fit and we expect it to be proportional to
the velocity dispersion of the progenitor, although this
proportionality needs to be checked more carefully (but see Figure 3
in \citealt{Sanders13a}, which shows that the size of the frequency
distribution scales as mass$^{1/3}$, or approximately as the velocity
dispersion). Similarly, the initial angle distribution, modeled here
as an isotropic Gaussian, has a charactaristic spread $\sigma_{\theta}$
which we set here to $0.003$ ($=\sigv/[122\kms]$), and we likewise
expect this spread to scale with the velocity dispersion of the
progenitor.

The fiducial model for a tidal stream in frequency--angle coordinates
proposed here is therefore described by essentially two free
parameters in addition to the progenitor's phase-space position: the
progenitor's velocity dispersion \sigv\ and the disruption time
$t_d$. As shown below, these two parameters are the most important in
determining the track of the stream. More sophisticated models can
leave $\mu_\Omega$ and $\sigma_{\theta}$ as free parameters as
well. In what follows, we will fix these to the values given in the
previous paragraph. It must be stressed, however, that all of these
parameters here have been fixed ``by-eye'', and better fits might be
possible using more quantitative fit procedures.

@@

In the generative model described above, the direction of the angle
along the stream lies along the principal eigenvector of $V_\veco$,
the variance matrix of the model $\Delta \veco$ distribution. We
denote the angle in this direction as $\apar$ and the frequency in
this direction $\opar$; the two-dimensional angle- and frequency-space
perpendicular to this direction is denoted as $\aperp$ and $\operp$.

First, we compute the distribution of $\Delta \opar$, the frequency
offset between stream members and the progenitor given the angle
offset $\Delta \apar$ as follows (considering $\Delta \opar$ to be
positive)
\begin{equation}\label{eq:poparapargeneral}
\begin{split}
  p(\Delta \opar|\Delta \apar) & = \int \dd \ts\,p(\Delta \opar,\ts|\Delta \apar)\, \\
  & \propto \int \dd \ts \,p(\Delta \apar | \Delta \opar,\ts)\,p(\Delta \opar,\ts)\,,\\
  & \approx \int \dd \ts \,\delta(\Delta \apar - \Delta \opar\,\ts)\,p(\Delta \opar,\ts)\,,\\
  & = \frac{1}{\Delta \opar}\,p(\Delta \opar,\frac{\Delta \apar}{\Delta \opar})\,,
\end{split}
\end{equation}
where in the penultimate step we have approximated the initial angle
distribution as a delta function (as the initial angle offset is small
with respect to the final offset). This is a fully general
expression. For the fiducial stream model, this can be simplified to
\begin{equation}\label{eq:poparapar}
\begin{split}
  p(\Delta \opar|\Delta \apar) \propto &\ \frac{1}{\Delta \opar}\,p(\Delta \opar)\,,\qquad 0 < \frac{\Delta \apar}{\Delta \opar} < t_d\,.\\
  & 0\,,\qquad \mathrm{otherwise}\,.
\end{split}
\end{equation}
\Eqnname~(\ref{eq:poparapar}) shows why we chose to model
$p(\Delta \opar)$ as a Gaussian multiplied with $|\Delta
\opar|$. $p(\Delta \opar|\Delta \apar)$ is then a Gaussian with mean
$\Delta \Omega^m$ and variance $\sigma^2_{\Omega,1}$, for $\Delta
\opar > \Delta \apar / t_d$, and zero otherwise. The mean and variance
of such a Gaussian is straightforward to calculate in terms of error
functions. The top left panel of \figurename~\ref{fig:gd1_apar} shows
the distribution of $|\Delta \opar|$ versus $|\Delta \apar|$ (absolute
value in order to show the trailing and leading stream together) for
the simulated stream, as well as the mean and dispersion calculated
from the fiducial model. The dashed line shows $\Delta \apar = \Delta
\Omega^m \,t_d$, approximately the angle where the finite disruption
time starts to influence the mean $\Delta \opar$ of the stream. The
finite disruption time, $\approx 4.5\Gyr$ in this case, of a tidal
stream means that for stream members to have reached very large angle
differences with respect to the progenitor, they must have been
removed at large frequency differences. Around $\Delta \Omega^m
\,t_d$, average stream stars, \ie, those removed with the average
frequency offset, do not have a large enough offset to reach large
angle offsets. Therefore, even though the average frequency offset
does not change over time (see \figurename~\ref{fig:gd1_times}), the
average frequency offset, and therefore the average orbit, does change
with distance from the progenitor. It is in this sense that streams do
not follow a single orbit. This is discussed further in BOVY:REF.

In the fiducial model, the average $\Delta \operp$, and $\Delta
\aperp$ as a function of $\Delta \apar$ are zero. Therefore, the
average stream track as a function of $\Delta \apar$ is entirely
specified by the mean $\Delta \opar(\Delta \apar)$ in the parallel
direction and zero offsets in the perpendicular directions. This track
can be rotated into the $(R,\phi,Z)$ coordinate system using the
eigenvectors of $V_\veco$.
